---
title: "Model_Building"
author: "Vanessa Zepeda"
output: html_document
---
```{r}
library(fastDummies)
library(tidyverse)
library(xgboost)
library(caret)
library(lubridate)
```


```{r}
# load in the dataset
df_to_clean = read_csv("~/Statistics_Practicum_2025/Final_Project/Data_Processing/tornado_final_dataset.csv")
df_to_clean
```

# Cleaning the Data
```{r}
# check for null values
sum(is.na(df_to_clean)) # 24 total (in population and landcover columns)
df_to_clean[!complete.cases(df_to_clean), ] # most are in Hawaii/Caribbean
df_to_clean = na.omit(df_to_clean) # drop null rows, no more nulls!

# categorical data to dummy
df_to_clean = fastDummies::dummy_cols(df_to_clean,
                                      select_columns = "landcover_majority500",
                                      remove_selected_columns = TRUE,
                                      remove_first_dummy = TRUE)

# convert time variables into cyclical values
df_to_clean = df_to_clean %>%
  mutate(
    hour = hour(time),
    time_sin = sin(2 * pi * (hour/24)),
    time_cos = cos(2 * pi * (hour/24)),
    month_sin = sin(2 * pi * (mo/12)),
    month_cos = cos(2 * pi * (mo/12))
  ) %>% select(-time, -mo, -hour)

# landcover 
df = df_to_clean %>% select(-geometry, -om, -fips, -elat, -elon, -yr, -dy)
df
```
## Assessing Class Imbalance
what method to choose?
- over sampling
- under sampling
- class weights
While the model is already robust when it comes to class imbalance, it is good practice to address the issue. Some common methods of addressing class imbalance are over sampling, which takes samples from the minority class. Another is under sampling, which removes samples from the majority class. And lastly, another is adding class weights, which we can use as the model supports this. But how do we choose the best approach?

class weights dont fully address the problem, however the other methods we risk losing data. For example, if we oversample the minority set, we may introduce unecessary noise to the dataset, leading to an overfit of the minority class. On the otherhand, under sampling the majority class could lose to a loss of a lot of data (would need to match (or be close to) the smallest count), which is not ideal either. While incorporating class weights is not the most ideal, that can be said for any other type of method we choose
```{r}
# assess class imbalance 
table(df$inj_class)
prop.table(table(df$inj_class))
```

```{r}
# for replicabiltiy
set.seed(67)
trainIndex = createDataPartition(df$inj_class, p = 0.7, list = FALSE)

# split the data into test/train data sets
dfTrain = df[trainIndex,]
dfTest = df[-trainIndex,]

# extract predictor and target variable(s)
X_train = dfTrain %>% select(-inj_class)
y_train = dfTrain$inj_class

X_test = dfTest %>% select(-inj_class)  
y_test = dfTest$inj_class

# check dimensions
dim(X_train)
length(y_train)
```

# baseline model without adding weights
While this is not a trustworthy model, we can see that the shap values make sense for what we can deduce intuitively;
- tornadoes that are larger in width and length are likely to be more destructive, therefore cause more damage 

1st round: 
nrounds 100, max_depth 6, eta 0.3, gamma 0, colsample_bytree 0.8, min_child_weight 3, subsample 0.8
kappa = 0.3912575
 
2nd round: 
nrounds 200, max_depth 6, eta 0.1, gamma 0.1, colsample_bytree 0.8, min_child_weight 3, subsample 0.8
kappa = 0.3895329

3rd round: 
nrounds 200, max_depth 6, eta 0.1, gamma 0.1, colsample_bytree 0.8, min_child_weight 3, subsample 0.8
kappa = 0.3895329

4th round:
nrounds 200, max_depth 6, eta 0.1, gamma 0.1, colsample_bytree 0.8, min_child_weight 3, subsample 0.8
kappa = 0.3895329

```{r}
# Set up parameter grid
xgb_grid <- expand.grid(
  nrounds = c(50, 100),
  max_depth = c(3, 6),
  eta = c(0.05, 0.1, 0.3),
  gamma = c(0, 1),
  colsample_bytree = c(0.8),
  min_child_weight = c(1, 3),
  subsample = c(0.8)
)

# Cross-validation control
train_control = trainControl(
  method = "cv",
  number = 3,
  verboseIter = TRUE,
)

# Train NON-weighted model
set.seed(67)
xgb_tuned_nonweighted = train(
  x = as.matrix(X_train),
  y = as.factor(y_train),
  trControl = train_control,
  tuneGrid = xgb_grid,
  method = "xgbTree",
  metric = "Kappa"
)

# View best parameters
xgb_tuned_nonweighted
plot(xgb_tuned_nonweighted)
best_params_nonweighted = xgb_tuned_nonweighted$bestTune
best_params_nonweighted

```


```{r}
## set params first
params = list(objective = "multi:softmax",
              eval_metric = "mlogloss",
              num_class = 3,
              eta = best_params_nonweighted$eta,
              max_depth = best_params_nonweighted$max_depth,
              gamma = best_params_nonweighted$gamma,
              colsample_bytree = best_params_nonweighted$colsample_bytree,
              min_child_weight = best_params_nonweighted$min_child_weight,
              subsample = best_params_nonweighted$subsample
)

## create model
baseline_model = xgboost(data = as.matrix(X_train),
                         label = y_train,
                         params = params,
                         nrounds = best_params_nonweighted$nrounds,
                         verbose = 1)
```

```{r}
## features that push target to either severe class or weak class: 
# if shap value is <0 injury severity is likely to be not that severe;
# if shap value is >0 injury severity is likely to be mroe severe 
xgb.plot.shap(model = baseline_model,
              data = as.matrix(X_train),
              top_n =4)

## Model Performance
# train model
pred = predict(baseline_model, as.matrix(X_test))

#error 
mean(pred != y_test) # error is low, which is good BUT based on the accuracy score, which cannot be trusted

# confusion matrix 
cm_un = confusionMatrix(factor(pred), factor(y_test))
```

# model with weights added
```{r}
# compute class counts from the training set
class_counts = table(y_train)
class_counts

# make weights inversely proportional to frequency
max_count = max(class_counts)
class_weights = as.numeric(max_count / class_counts)
class_weights_named = setNames(class_weights, names(class_counts))

class_weights_named

# assign weights to each training row
train_weights = class_weights_named[ as.character(y_train) ]
train_weights
```

```{r}
# Set up parameter grid
xgb_grid = expand.grid(
  nrounds = c(50, 100),
  max_depth = c(3, 6),
  eta = c(0.05, 0.1, 0.3),
  gamma = c(0, 1),
  colsample_bytree = c(0.8),
  min_child_weight = c(1, 3),
  subsample = c(0.8)
)

# Cross-validation control
train_control = trainControl(
  method = "cv",
  number = 3,
  verboseIter = TRUE,
)

# Train weighted model
set.seed(67)
xgb_tuned_weighted = train(
  x = as.matrix(X_train),
  y = as.factor(y_train),
  weights = train_weights,
  trControl = train_control,
  tuneGrid = xgb_grid,
  method = "xgbTree",
  metric = "Kappa"
)
```

```{r}
# View best parameters
xgb_tuned_weighted
plot(xgb_tuned_weighted)
best_params_weighted = xgb_tuned_weighted$bestTune
best_params_weighted
```


```{r}
# fit weighted XGBoost model

## set params first
params = list(objective = "multi:softmax",
              eval_metric = "mlogloss",
              num_class = 3,
              eta = best_params_weighted$eta,
              max_depth = best_params_weighted$max_depth,
              gamma = best_params_weighted$gamma,
              colsample_bytree = best_params_weighted$colsample_bytree,
              min_child_weight = best_params_weighted$min_child_weight,
              subsample = best_params_weighted$subsample
)

## create model
weighted_model = xgboost(data = as.matrix(X_train),
                         label = y_train,
                         params = params,
                         nrounds = best_params_weighted$nrounds,
                         verbose = 1)

```

```{r}
# SHAP plot for weighted model
xgb.plot.shap(
model = weighted_model,
data = as.matrix(X_train),
top_n = 4,
n_col = 2
)

# Evaluate performance
pred_weighted = predict(weighted_model, as.matrix(X_test))

# confusion matrix
cm = confusionMatrix(
  factor(pred_weighted),
  factor(y_test)
)

# misclassification rate
mean(pred_weighted != y_test)
```
```{r}

precision <- cm_un$byClass[,"Pos Pred Value"]
recall    <- cm_un$byClass[,"Sensitivity"]

F1 <- 2 * (precision * recall) / (precision + recall)
F1_macro <- mean(F1, na.rm = TRUE)

F1
F1_macro
```
```{r}
cm_df = as.data.frame(cm$table)
colnames(cm_df) = c("Predicted", "Actual", "Freq")

ggplot(cm_df, aes(x = Actual, y = Predicted, fill = Freq)) +
  geom_tile() +
  scale_fill_gradient(low = "lightblue", high = "steelblue") + 
  geom_text(aes(label = Freq)) +
  labs(title = "Weighted Model Confusion Matrix Heatmap") 
```

```{r}

xgb.plot.shap(
  model = weighted_model,
  data = as.matrix(X_train),
  top_n = 4,
  n_col = 2
)
```


```{r}
ggplot(df, aes(x = factor(inj_class))) +
  geom_bar(fill = "steelblue") +
  labs(title = "Class Distribution of Injury Severity",
       x = "Injury Class", y = "Count")

```

```{r}
importance_matrix = xgb.importance(
  model = weighted_model,
  feature_names = colnames(X_train)
)

xgb.plot.importance(importance_matrix, top_n = 20)
```

```{r}
df2 = df %>% mutate(hour = (atan2(time_sin, time_cos) * 24/(2*pi)) %% 24)

ggplot(df2, aes(x = factor(inj_class), y = hour, fill = factor(inj_class))) +
  geom_boxplot() +
  labs(title = "Distribution of Tornado Occurrence Hour by Injury Severity",
       x = "Injury Class", y = "Hour of Day")
```


```{r}
df2 = df %>% mutate(month = (atan2(month_sin, month_cos) * 12/(2*pi)) %% 12)

ggplot(df2, aes(x = month, fill = factor(inj_class))) +
  geom_density(alpha = 0.5) +
  labs(title = "Seasonal Pattern by Injury Severity",
       x = "Month", y = "Density")
```

```{r}

ggplot(dfTest, aes(x = slon, y = slat, color = factor(pred_weighted))) +
  geom_point(alpha = 0.4) +
  labs(title = "Predicted Injury Severity (Spatial Scatter)",
       x = "Longitude", y = "Latitude")
```

